Investigating The Fleishman Effect: Subjective Evaluations of Financial Satisfaction in the US General Social Survey

### 2023-03-14 outputs/paper/financial_satsifaction_in_us.qmd
edited discussion

| Variable | New Name               | Description                                                                             | Example               |
|----------------|----------------|------------------------|----------------|
| satfin   | financial_satisfaction | Respondent's satisfaction with their current financial situation                        | Pretty Well Satisfied |
| finalter | financial_change       | Changes in respondent's financial situation                                             | Stayed the same       |
| finrela  | financial_compare      | Comparing respondent's family income with American families in general                  | Above Average         |
| class    | social_class           | self identified social class the respondent belong in                                   | Middle Class          |
| class1   | social_class1          | self identified social class the respondent belong in (more granular than social_class) | Upper Middle Class    |

@@

Our results point to three trends: first, our visualizations illustrate a growing wealth gap over time. Second, the data suggests that financial well-being and satisfaction is influenced by demographic factors. And third, evidence to support the Fleishman effect \-- the idea that upper classes have a distorted sense of their financial well-being, since they compare themselves to the upper class, rather than comparing themselves to the general population.

## Growing Wealth Gap Impacts Demographic Groups Differently

Our analysis shows that the gap between high-earning and low-earning individuals in the US has been increasing since 1970, when the GSS was first recorded. Even though GDP is rising, we don't observe a significant improvement in financial satisfaction over time. In fact, twice as many respondents identify as "lower class" in 2021 than in 1970.

Respondents' gender (defined in the GSS as sex at birth) and level of education have a significant influence on financial satisfaction. Men tend to see their financial satisfaction as improving, and report greater levels of satisfaction than women (@fig-2021-gender-financial). Respondents with Bachelors or Graduate degrees are more satisfied, and describe their financial situations as getting better over time (@fig-2021-degree-financial) . This suggests that the increased financial dissatisfaction reported over time in @fig-overtime-1 is more likely to be felt by non-male individuals with less formal education.

In addition to gender and education, generational cohorts are impacted differently by the growing wealth gap. In @fig-overtime-cohort, the Greatest, Silent, and Boomer generations all see an increase in individuals reporting "Pretty Well Satisfied" over the course of their lives. In Gen X, this "Pretty Well Satisfied" line flattens, and Millennials and Gen Z see a significant decrease in high satisfaction (of course, Gen Z's visualization is correlated with a global pandemic).

While satisfaction can often be correlated with income -- for instance, in @fig-2021-income-financial those who are making lower income are less satisfied-- it shouldn't be confused with income. The way somebody feels about the financial situation might not reflect the reality of their financial situation as held against some other objective measure. In @fig-overtime, we see that reported financial situations are either getting better or getting worse over time, but that dissatisfaction overall is rising (with nearly 20% of upper-class respondents reporting dissatisfaction in @fig-socialclass-financial). Although the levels of respondents identifying as upper and middle classes has remained relatively stable over time (@fig-socialclass-overtime), there is a general decline is people reporting being "Pretty Well Satisfied", and an increase in individuals reporting "Not Satisfied At All" (@fig-overtime). Who is feeling this dissatisfaction? Why?

@@

Our data on financial satisfaction based on income supports this hypothesis (@fig-socialclass-compare). When faceted by class instead, the question becomes more interesting. When asked to compare themselves to the general population, nearly 20% of self-identified upper-classes see themselves as "below average". While this may be due to problems with the survey recording, or respondents misunderstanding the question, the fact that only 60% of upper class individuals see themselves as being far above average suggests some sort of distortion of self-perception once that level of wealth is achieved.

Is the upper class simply delusional? Could survey respondents have misinterpreted what "upper class" means? Or, are there more factors at play when people consider their own financial situations \-- for instance, the child of millionaires who themselves make \$40k a year might see themselves as culturally and socially upper class, yet below average in terms of financial comparison.

In her 2019 memoir We Need To Talk About Money [@talkAboutMoney], Otegha Uwagba writes about coming of age as a Black, Oxford-educated millennial woman from a low-income family, and how her perception of her own financial well-being was distorted by the way her upper class colleagues were able to afford luxuries like property ownership while signalling, by complaining about mortgages and budgeting, that they were in the same cohort of struggling millennials. After learning that a friend had concealed that she, in fact, didn't have a mortgage, and that her parents had given her enough money to buy her apartment outright, Uwagba writes about the cultural shame that politically liberal high-earners (or inheritors) can experience as anti-capitalist sentiments have entered the mainstream conversations:

"But that's the thing about shame -- it's an emotion that arises from our most base human instincts: the fear of being judged negatively by others; our desire to 'fit in'; the discomfort of feeling exposed; and those emotional responses transcend income levels. For wealthy people who also happen to be politically liberal -- and so value the notion of social equity -- being rich, and the advantages that wealth presents can be a source of moral discomfort, particularly when said wealth is inherited."

Concealing wealth out of shame, to Uwagba, is "actively unhelpful" for others in the same reference group, and exacerbates the frustration of realizing that achieving personal finance goals may even more out of reach than initially believed. In @fig-overtime, we see that, beginning with Gen X, Gen Xers, Millennials, and Gen Z are experiencing more financial dissatisfaction over time compared to other generations (although, Gen Z's self-reporting does correlate with a pandemic). This complicates this idea of reference group comparisons \-- while intra-group comparisons are full of cognitive biases and emotions such as shame that distort people's understandings of their own and their colleagues financial well-being, younger generations can no longer use their own age and educational achievements as a reference group. Millennials aren't experiencing the same financial satisfaction that Baby Boomers did when they were the same age.

Meanwhile, the lure of upward mobility is addictive, as even the upper class reports financial dissatisfaction. An article by The Cut titled "The Fleishman Effect" [@fleishmanEffect] explores how upper-class, New York women are responding to the HBO show *Fleishman Is In Trouble*, which is about the stress that high-earning women feel when trying to achieve satisfaction in a city that houses one of the wealthiest reference groups in the world, where "the go-to bat mitzvah gift of the moment is a Cartier bracelet, for which moms are expected to pitch in for a group present". When held against some objective measure of financial well-being, being part of a group that gifts a child with designer jewelry suggests an income and lifestyle that would suggest financial satisfaction. "The Fleishman Effect" occurs when these higher group standards push satisfaction out of reach.

@@

Financial satisfaction is dependent on an individual's context, and much more subjective than a fixed measure like individual or household income. We hypothesize that this context may be determined not only by their current reference group, but also by their relationship to wealth, and how this relationship evolved over the course of their life. The study would be strengthened with data that could provide a fuller profile of respondent's reference groups, and other factors that influence their motivations and attitudes towards personal finance and upward mobility.

The way that individuals narrativize their success in comparison to their upbringing may influence the way that they interpret their financial standing. In our supplemental survey in Appendix X, we suggest the questions "Have you moved up a full socio-economic class compared to your upbringing? (Y/N/Don't Know", "Do you have at least one parent who moved up a full socio-economic class?" (Y/N/Don't Know), and "Is your current career in the same field as your parents?" (Y/N/Don't Know) to better understand respondent's family histories. Asking "Are your friends in the same socio-economic class, higher, or lower? (Higher/About the same/Lower)" would also provide more granular detail into how someone perceives their financial status in comparison to the people closest to them. The Social Capital Atlas [@socialCapital] conducted a survey pursuing a similar estimand using Facebook data to explore "economic connectedness" in friend groups, which refers to the percentage of friends a respondent has who earn higher or lower incomes than them.

@fig-2021-degree-financial shows that respondents with post-secondary education experience better changes in financial situation, with 50% of respondents with graduate degrees reporting that their financial situation is "getting better". Obtaining more details on these reference groups by their area of study, and whether or not they regret their major, could provide more information on how the different returns-on-investment on different degrees change over time, as shown by The Washington Post's reporting on the Federal Reserve Survey of Household Economics and Decision Making [@collegeMajors].

We propose adding the questions "If you went to college or university, what was your major?" "Do you regret your major? (Y/N/Don't know)" to the survey. Adding this question to our study would provide more detailed insight on reference groups, since we could see how people who enrolled in each major compare themselves to others, and, with more information on family history, we could also gather insight on which fields individuals from different social classes pursue in higher education. We could also see how people across different income ranges and classes experience regret about their decisions.

Geography can have a major influence on personal finance and reference groups. A respondent making \$75,000 a year living in Hattiesburg, Mississippi might feel more financially satisfied than a respondent with the same income living in San Francisco, California.  However, this could sacrifice anonymity. This question would still have to be broad enough to provide information on their location (i.e. by city), rather than something that could de-anonymize respondents when triangulated with other data (i.e. postal code).

@@

In conclusion, financial satisfaction and well-being are subjective and influenced by various factors. This paper used data from the US General Social Survey to examine how people perceive their financial situations in comparison to others, particularly in light of the increasing wealth gap between high and low earning individuals.

The results support the hypothesis that the rich are getting richer, the poor are getting poorer, and that younger generations are financially worse off than preceding cohorts. The data also suggests that people's reference groups influence their understanding of their financial well-being and satisfaction. Overall, this paper highlights the complexity of financial satisfaction and well-being, and the importance of considering sociological dynamics and cognitive biases in understanding these subjective evaluations.

>>>>

| Variable | New Name               | Description                                                                             | Example               |
|-----------------|-----------------|----------------------|-----------------|
| satfin   | financial_satisfaction | Respondent's satisfaction with their current financial situation                        | Pretty Well Satisfied |
| finalter | financial_change       | Changes in respondent's financial situation                                             | Stayed the same       |
| finrela  | financial_compare      | Comparing respondent's family income with American families in general                  | Above Average         |
| class    | social_class           | self identified social class the respondent belong in                                   | Middle Class          |
| class1   | social_class1          | self identified social class the respondent belong in (more granular than social_class) | Upper Middle Class    |

@@

Our results point to three trends: first, our visualizations illustrate a growing wealth gap over time. Even though GDP is rising, we don't observe a significant improvement in financial satisfaction over time. In fact, twice as many respondents identify as "lower class" in 2021 than in 1970. Second, the data suggests that financial well-being and satisfaction is influenced by demographic factors. And third, evidence to support the Fleishman Effect \-- a term that refers to self-identified upper class respondents who have a distorted sense of their financial well-being, since they compare themselves to other elites.

## Demographic Factors' Influence on Financial Satisfaction

According to the findings of our study, the impact of macro-economic trends like the widening wealth gap on individuals varies based on their demographic characteristics. Specifically, women and those with less formal education tend to report increased dissatisfaction over time. Men tend to see their financial satisfaction as improving, and report greater levels of satisfaction than women (@fig-2021-gender-financial). Respondents with Bachelors or Graduate degrees are more satisfied, and describe their financial situations as getting better over time (@fig-2021-degree-financial) .

In addition to gender and education, generational cohorts are impacted differently by the growing wealth gap. In @fig-overtime-cohort, the Greatest, Silent, and Boomer generations all see an increase in individuals reporting "Pretty Well Satisfied" over the course of their lives. In Gen X, this "Pretty Well Satisfied" line flattens, and Millennials and Gen Z see a significant decrease in high satisfaction (of course, Gen Z's visualization is correlated with a global pandemic, and represents respondents at the beginning of their careers).

While satisfaction can often be correlated with income -- for instance, in @fig-2021-income-financial those who are making lower income are less satisfied-- it shouldn't be confused with income. The way somebody feels about the financial situation might not reflect the reality of their financial situation as held against some other objective measure. For instance, nearly 20% of upper-class respondents report dissatisfaction in @fig-socialclass-financial. In @fig-overtime, we see that reported financial situations are either getting better or getting worse over time, but that dissatisfaction overall is rising. Although the levels of respondents identifying as upper and middle classes has remained relatively stable over time (@fig-socialclass-overtime), there is a general decline is people reporting being "Pretty Well Satisfied", and an increase in individuals reporting "Not Satisfied At All" (@fig-overtime). Who is feeling all of this dissatisfaction? Why?

@@

Our data on financial satisfaction based on income supports this hypothesis. When asked to compare themselves to the general population, nearly 20% of self-identified upper-classes see themselves as "below average" (@fig-socialclass-compare). While this may be due to problems with the survey recording, or respondents misinterpreting the question, the fact that only 60% of upper class individuals see themselves as being far above average suggests some sort of distortion of self-perception once that level of wealth is achieved.

Is the upper class simply delusional? Could survey respondents have misinterpreted what "upper class" means? Or, are there more factors at play when people consider their own financial situations \-- for instance, the child of millionaires who themselves make \$40k a year might see themselves as culturally and socially upper class, yet below average in terms of income.

In her 2019 memoir *We Need To Talk About Money* [@talkAboutMoney], Otegha Uwagba writes about coming of age as a Black, Oxford-educated millennial woman from a low-income family, and how her perception of her own financial well-being was distorted by the way her upper class colleagues were able to afford luxuries like property ownership while signalling, by complaining about mortgages and budgeting, that they were in the same cohort of hustling milennials struggling to achieve the same checkpoints of financial success as previous generations. After learning that a friend had concealed that she, in fact, didn't have a mortgage, and that her parents had given her enough money to buy her apartment outright, Uwagba writes about the cultural shame that politically liberal high-earners (or inheritors) can experience as anti-capitalist sentiments have entered the mainstream conversations:

> But that's the thing about shame -- it's an emotion that arises from our most base human instincts: the fear of being judged negatively by others; our desire to 'fit in'; the discomfort of feeling exposed; and those emotional responses transcend income levels. For wealthy people who also happen to be politically liberal -- and so value the notion of social equity -- being rich, and the advantages that wealth presents can be a source of moral discomfort, particularly when said wealth is inherited.

Concealing wealth out of shame, to Uwagba, is "actively unhelpful" for others in the same reference group, and exacerbates the frustration of realizing that achieving personal finance goals may even more out of reach than initially believed. In @fig-overtime, we see that, beginning with Gen X, Gen Xers, Millennials, and Gen Z are experiencing more financial dissatisfaction over time compared to other generations (although, Gen Z's self-reporting does correlate with a pandemic). This complicates this idea of reference group comparisons \-- while intra-group comparisons are full of cognitive biases and emotions such as shame that may distort people's understandings of their own and their colleagues financial well-being, younger generations can no longer use their own age and educational achievements as a reference group. Millennials aren't experiencing the same financial satisfaction that Baby Boomers did when they were the same age.

Some upper-class individuals, specifically those belonging to generational cohorts experiencing financial distress, might therefore be persuade to perform and internalize a level of financial dissatisfaction that doesn't actually correspond to their wealth. On the other hand, financial dissatisfaction in the upper-classes can also come from a reference group effect, where their frame of comparison is no longer the average American.

An article by The Cut titled "The Fleishman Effect" [@fleishmanEffect] explores how upper-class, New York women are responding to the FX television show *Fleishman Is In Trouble*, which is about, among other things, the pressures that high-earning women feel when trying to achieve satisfaction in a city that houses one of the wealthiest reference groups in the world, where "the go-to bat mitzvah gift of the moment is a Cartier bracelet, for which moms are expected to pitch in for a group present". When held against some objective measure of financial well-being, being part of a group that gifts a child with designer jewelry suggests an income and lifestyle that would suggest financial satisfaction. "The Fleishman Effect" occurs when these higher group standards push satisfaction out of reach.

While the GSS data cannot provide a full enough profile of its respondents to determine the exact social and cultural conditions that their perceptions are shaped by, it does suggest that financial satisfaction is an individualized and subjective experience, and that demographic categories and class dynamics have an impact on responses.

@@

Financial satisfaction is dependent on an individual's context, and much more subjective than a fixed measure like individual or household income. We hypothesize that this context may be determined not only by their current reference group, but also by their relationship to wealth, and how this relationship evolved over the course of their life. The study would be strengthened by data that can provide a fuller profile of respondent's reference groups, and other factors that influence their motivations and attitudes towards personal finance and upward mobility.

We hypothesize that the way that individuals narrativize their success in comparison to their upbringing may influence the way that they interpret their financial standing. In our supplemental survey in the Appendix, we suggest the questions "Have you moved up a full socio-economic class compared to your upbringing? (Y/N/Don't Know", "Do you have at least one parent who moved up a full socio-economic class?" (Y/N/Don't Know), and "Is your current career in the same field as your parents?" (Y/N/Don't Know) to better understand respondent's family histories. Asking "Are your friends in the same socio-economic class, higher, or lower? (Higher/About the same/Lower)" would also provide more granular detail into how someone perceives their financial status in comparison to the people closest to them. The Social Capital Atlas [@socialCapital] conducted a survey pursuing a similar estimand using Facebook data to explore "economic connectedness" in friend groups, which refers to the percentage of friends a respondent has who earn higher or lower incomes than them.

@fig-2021-degree-financial shows that respondents with post-secondary education experience more positive changes in financial situation, with 50% of respondents with graduate degrees reporting that their financial situation is "getting better". Obtaining more detail on these respondents' areas of study, and whether or not they regret their major, could provide more information on how the returns-on-investment of different degrees change over time. This data can be cross-referenced with *The Washington Post*'s analysis on which majors are the most regretted, which draws on data from Federal Reserve Survey of Household Economics and Decision Making [@collegeMajors].

We propose adding the questions "If you went to college or university, what was your major?" "Do you regret your major? (Y/N/Don't know)" to the survey. Adding this question to our study would provide more detailed insight on reference groups, since we could see how people who enrolled in each major compare themselves to others, and, with the additional information on their family's class history, we could also gather insight on which fields individuals from different social classes might pursue in higher education. We could also see how people across different income ranges and classes experience regret about their decisions.

Geography can have a major influence on personal finance and reference groups. A respondent making \$75,000 a year living in Hattiesburg, Mississippi might feel more financially satisfied than a respondent with the same income living in San Francisco, California. However, collecting data on respondents' location could sacrifice anonymity. This question would still have to be broad enough to provide information on their location (i.e. by city), rather than something that could de-anonymize respondents when triangulated with other data (i.e. postal code).

@@

This paper used data from the US General Social Survey to examine how people perceive their financial situations in comparison to others, particularly in light of the increasing wealth gap between high and low earning individuals. We found that significant trends in financial satisfaction and comparison based on demographic categories such as generational cohort, level of education, age, gender, and social class.The results support the hypothesis that the rich are getting richer (or at least, feel like it), the poor are getting poorer (or at least, feel like it), and that younger generations are financially worse off than preceding cohorts. The data also suggests that people's reference groups influence their understanding of their financial well-being and satisfaction, with upper-classes experiencing counter-intuitively significant rates of financial dissatisfaction. Overall, this paper highlights the complexity of financial satisfaction and well-being, and the importance of considering sociological dynamics and cognitive biases in understanding these subjective evaluations. Further research with additional data on factors that shape a respondents' self-narrative, such as upbringing, family histories of social mobility, or the areas of study pursued in post-secondary education could add to our understandings of reference groups as they relate to wealth and perceived success.
